A lottery model of density-dependent selection in evolutionary genetics

### 2017-05-05 manuscript.tex
latest

Using our extended lottery model, we then revisit Grime's C/S/R scheme, and evaluate the extent to which selection favors fecundity, competitive ability or lower mortality under Grime's environmental extremes (section ``Primary strategies and Grime's triangle''). This represents a simple ``sanity check'' on the invasion behavior of our extended lottery model in different extreme cases.

@@

The resulting formulation of the C/S/R scheme is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g. \citealt{tilman_2007}).

Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}. In addition, we have not assumed any kind of trade-offs or pleiotropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$. As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances \citep{darling_2012}. Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time. Thus, while trade-offs may amplify specialization, and are sometimes invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}, they are not necessary for it.

>>>>

Using our extended lottery model, we then revisit Grime's C/S/R scheme, and evaluate the extent to which selection favors fecundity, competitive ability or lower mortality under Grime's environments (section ``Primary strategies and Grime's triangle''). This represents a simple ``sanity check'' on the invasion behavior of our extended lottery model under different extremes.

@@

Our realization of Grime's triangle (Fig.~\ref{fig:axes}) differs from approaches which identify primary strategies as trait combinations which can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}. Moreover, our formulation is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g. \citealt{tilman_2007}). However, section ``Primary strategies and Grime's triangle'' suggests that, apart from the obvious irrelevance of territorial contests at low density, selection alone has no preference for fecundity, competitive ability or longevity. This could mean that one or both of the factors we set aside --- pleiotropy/trade-offs and mutation bias --- are needed to get  true ``primary strategy'' trait differentiation. A separate model of trade-offs and mutation effects would be needed to explore this possibility.

Nevertheless, it is interesting to note that ruderals, which are typically thought of as high fecundity dispersers ($b$ specialists), may also be strongly $d$-selected, consistent with our findings. An effective way to reduce $d$ in the face of unavoidable physical destruction is to shorten the time to reproductive maturity --- short life cycles are a characteristically ruderal trait. Moreover, a recent hierarchical cluster analysis of coral traits did find a distinct ``ruderal'' cluster, but high fecundity was not its distinguishing feature. Rather, ruderals used brood- (as opposed to broadcast-) spawning, which could plausibly be a mechanism for improving propagule survivorship in disturbed environments \citep{darling_2012}.
